A Plasma Torus Around a Young Low-Mass Star

### 2025-03-14 paper/ms.tex
update

\begin{abstract}
\normalfont
Approximately one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material corotating with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
starspots or faculae distributed across the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a young ($t$=16\,Myr), rapidly-rotating ($P$=3.9\,hr)
CPV, TIC~141146667. The spectra show coherent sinusoidal Balmer
emission at twice to four times the star's equatorial velocity,
demonstrating the presence of extended clumps of circumstellar
plasma --- a plasma torus.  Given that long-lived condensations of
cool ($10^4$ K) plasma can persist in the hot ($10^6$ K) coronae of
stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can be optically thick; the exact origin of the
material and its microphysical opacity remain unclear.
\end{abstract}

@@

$M_\star$\dotfill                       & Stellar mass ($M_\odot$)\dotfill                               & 0.19$\pm$0.01  & 3 \\

@@

{\it Effective temperature, radius, mass, and spectral type}
We determined the stellar effective temperature and radius by adopting
the results from the spectral energy distribution fitting procedure
described by \cite{Bouma2024}.  To summarize, this approach used
\texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we used the broadband magnitudes from Gaia DR2, APASS,
2MASS, SDSS, and WISE $W1$ and $W2$.  This method has the most
constraining power for the star's effective temperature (2972 $\pm$
40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface gravity,
metallicity, and reddening are typically only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
against the photometric standards presented by \cite{Bochanski2007};
this qualitatively agrees with the effective temperature from the SED
fitting \cite{Pecaut2013}.  With the effective temperature, stellar
radius, and previously determined age, we then estimated the stellar
mass by interpolating against the PARSEC v1.2S isochrones
\cite{Chen2014}.  Specifically, we used the distance metric defined in
Equation~8 of \cite{Bouma2024} to select the model mass closest to the
observed temperature, radius, and age.  We derived the uncertainty in
the mass based by propagating the statistical uncertainties in each of
these parameters.

@@

{\it Age: Isochrones}---The color and absolute magnitude of
TIC~141146667 suggest that it is a pre-main sequence M dwarf, similar
to all other known CPVs \cite{Stauffer2017,Stauffer2021,Bouma2024}.
The star's proximity ($d$=58\,pc) and its high galactic latitude
($b$=$+$53$^\circ$) yield negligible interstellar reddening along the
line of sight \cite{Green2019}.  Figure~\ref{fig:camd} shows the
location of the star in the color--absolute magnitude diagram (CAMD)
relative to young stellar populations including Upper Scorpius (USco),
IC~2602, and the Pleiades.  To make this diagram, we adopted the USco
members in the $\delta$~Sco and $\sigma$~Sco sub-associations from
\cite{Ratzenbock2023}, and the IC~2602 and Pleiades members from
\cite{Hunt2024}.  We assumed an average $V$-band extinction $A_{\rm
V}$=${0.12, 0.11, 0.10}\,{\rm mag}$ for USco \cite{Pecaut2016},
IC~2602 \cite{Hunt2024}, and the Pleiades \cite{Hunt2024}
respectively.  We dereddened the photometry using the extinction
coefficients $k_X\equiv A_X/A_0$ tabulated in
\citet{GaiaCollaboration2018}, assuming that $A_0 = 3.1 E(B-V)$.

Figure~\ref{fig:camd} suggests that TIC~141146667 is
older than USco, and potentially younger than the Pleiades.  The precision of this
statement is set by the intrinsic scatter of the calibration
sequences; the most luminous stars in the Pleiades of the same color
have a similar absolute magnitude as TIC~141146667.
Previous work \cite{Stauffer2021} has noted that in the Gaia
passbands, CPVs tend to be photometrically redder and more luminous
than single stars in any given cluster.
While this effect complicates any attempt at age inference based on this
photometry, it suggests that the Pleiades may be a better comparison
population than IC~2602.
Broadly, we take the star's isochrone location to suggest age bounds
$t_{\rm CAMD}$$\sim$40-150\,Myr.

@@

Desite uncertainties in the details, what can be stated with
confidence is that lithium is not abundant in the spectrum of
TIC~141146667.  Figure~\ref{fig:liew_population} compares our upper
limit against the lithium equivalent width measurements reported by
\cite{Jeffries2023} based on the Gaia-ESO spectroscopic survey.  If
the star were $\lesssim$20\,Myr old, given its temperature we would
expect to see lithium in abundance ($>$400\,m\AA).  Since we do not,
we conclude that $t_{\rm Li,emp}$$\gtrsim$20\,Myr.  The
\cite{Feiden2016} isochrones provide a point for theoretical
comparison, suggesting that since $M_{\rm K}$$=$6.67\,mag, $t_{\rm
Li,th}$$\gtrsim$35\,Myr would be expected for complete depletion (see
e.g.~Figure 7 from \cite{Wood2023}).

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf3.pdf}
  \caption{Dereddened Gaia DR3 color vs.~absolute magnitude diagram of
  TIC~141146667 and comparison samples. 
  TIC~141146667 is on the pre-main sequence; stars with the same color
  that have arrived at the main sequence are typically 1.5\,magnitudes
  fainter.  The star's location in this diagram broadly suggests an
  age of 40-110\,Myr.  }
  \label{fig:camd}
\end{figure}

@@

{\it Age: Summary}---The main indicators for the youth of
TIC~141146667 are {\it i)} that it is a complex periodic variable, and
{\it ii)} that it is 1.5 magnitudes brighter than main sequence stars
of the same color, while showing no indicators for binarity.  The
first item constitutes evidence for youth since all known CPVs are
$\lesssim$200\,Myr \cite{Bouma2024}.  The reference sequences for the
Pleiades (112\,Myr, \cite{Dahm2015}) show a few stars of equal
luminosity and the same temperature, suggesting a photometric
isochronal age upper limit $\lesssim$150\,Myr.  The weak lithium
absorption suggests an age of at least 20\,Myr based on an empirical
comparison using Gaia-ESO spectra, or at least 35\,Myr based on the
\cite{Feiden2016} isochrones.  These considerations yield our adopted
age range of 35-150\,Myr.

>>>>

\begin{abstract}
\normalfont
Approximately one percent of red dwarfs younger than 100 million years
show structured, periodic optical light curves suggestive of
transiting opaque material corotating with the star
\cite{Rebull2016,Stauffer2017,Rebull2018,Bouma2024}.  The
composition, origin, and even the existence of this material are
uncertain. The main alternative hypothesis is that these complex
periodic variables (CPVs) are explained by complex distributions of
starspots or faculae distributed across the stellar surfaces
\cite{Koen2021}.  Here, we present time-series spectroscopy and
photometry of a rapidly-rotating ($P$=3.9\,hr)
CPV, TIC~141146667. The spectra show coherent sinusoidal Balmer
emission at twice to four times the star's equatorial velocity,
demonstrating the presence of extended clumps of circumstellar
plasma --- a plasma torus.  Given that long-lived condensations of
cool ($10^4$ K) plasma can persist in the hot ($10^6$ K) coronae of
stars with a wide range of masses
\cite{CollierCameron1989,Townsend2005,Dunstone2006,Petit2013,Waugh2022,Daley-Yates2024},
these data support the idea that around the lowest-mass stars, such
condensations can be optically thick; the exact origin of the
material and its microphysical opacity remain unclear.
\end{abstract}

@@

$M_\star$\dotfill                       & Stellar mass ($M_\odot$)\dotfill                               & 0.22$\pm$0.02  & 3 \\

@@

{\it Effective temperature, radius, mass, and spectral type}
We determined the stellar effective temperature and radius by adopting
the results from the spectral energy distribution fitting procedure
described by \cite{Bouma2024}.  To summarize, this approach used
\texttt{astroARIADNE} \cite{Vines2022} with the BT-Settl stellar
atmosphere models \cite{Allard2012}, assuming the \cite{Asplund2009}
solar abundances and the \cite{Barber2006} water line lists.  This
approach fitted for the stellar effective temperature, radius,
reddening, surface gravity, metallicity, and distance by comparing the
measured broadband magnitudes against pre-computed model grids.
Specifically, we used the broadband magnitudes from Gaia DR2, APASS,
2MASS, SDSS, and WISE $W1$ and $W2$.  This method has the most
constraining power for the star's effective temperature (2972 $\pm$
40\,K) and radius (0.42 $\pm$ 0.02\,$R_\odot$); the surface gravity,
metallicity, and reddening are typically only weakly constrained.  We
determined the star's spectral type to be M5.5Ve by visually comparing
against the photometric standards tabulated by \cite{Bochanski2007},
which agrees with the effective temperature from the SED fitting
\cite{Pecaut2013}.  Given the effective temperature, stellar radius, and
age range (35-150\,Myr) derived below, we then estimated the stellar
mass by interpolating against the PARSEC v1.2S isochrones
\cite{Chen2014}.  Specifically, we used the distance metric defined in
Equation~8 of \cite{Bouma2024} to select the model mass closest to a
given observed temperature, radius, and age.  This exercise yielded a
mass of $M_\star$=$0.20^{+0.02}_{-0.01}$ assuming an age of 35\,Myr, or
a mass of $0.25^{+0.02}_{-0.01}$ assuming an age of 150\,Myr.  
These in turn imply Keplerian corotation radii $R_{\rm
cr}/R_\star$=1.75$\pm$0.07 and $R_{\rm cr}/R_\star$=1.89$\pm$0.07,
respectively; this particular scale is relevant as it is theoretically
expected to set the inner boundary at which corotating material can
accumulate (e.g.~\cite{Townsend2005,Daley-Yates2024}).
For our final quoted $M_\star$ and $R_{\rm cr}$ values we take a
geometric mean of the extremes; we caution however that this choice is
arbitrary, and that a more precise age would be needed to circumvent
the systematic uncertainties in these parameters.

@@

{\it Age: Isochrones}---The color and absolute magnitude of
TIC~141146667 suggest that it is a pre-main sequence M dwarf, similar
to all other known CPVs \cite{Stauffer2017,Stauffer2021,Bouma2024}.
The star's proximity ($d$=58\,pc) and its high galactic latitude
($b$=$+$53$^\circ$) yield negligible interstellar reddening along the
line of sight \cite{Green2019}.  Figure~\ref{fig:camd} shows the
location of the star in the color--absolute magnitude diagram (CAMD)
relative to young stellar populations including Upper Scorpius (USco),
IC~2602, and the Pleiades.  To make this diagram, we adopted the USco
members in the $\delta$~Sco and $\sigma$~Sco sub-associations from
\cite{Ratzenbock2023}, and the IC~2602 and Pleiades members from
\cite{Hunt2024}.  We assumed an average $V$-band extinction $A_{\rm
V}$=$\{0.12, 0.11, 0.10\}\,{\rm mag}$ for USco \cite{Pecaut2016},
IC~2602 \cite{Hunt2024}, and the Pleiades \cite{Hunt2024}
respectively.  We dereddened the photometry using the extinction
coefficients $k_X\equiv A_X/A_0$ tabulated in
\cite{GaiaCollaboration2018}, assuming that $A_0 = 3.1 E(B-V)$.

Figure~\ref{fig:camd} shows that TIC~141146667 falls between
the USco and Pleiades sequences, while nearly overlapping with
IC~2602.  However, the precision of the implied age
is set by the intrinsic scatter of these calibration
sequences; the most luminous stars in the Pleiades of the same color
have a similar absolute magnitude as TIC~141146667.
Previous work \cite{Stauffer2021} has also noted that in the Gaia
passbands, CPVs tend to be photometrically redder and more luminous
than single stars in any given cluster, similar to other rapid
rotators.
While this effect complicates any attempt at age inference based on
the Gaia photometry, it suggests that the Pleiades may be a better
comparison population than IC~2602.  Broadly, we take the star's
isochrone location to suggest age bounds $t_{\rm
CAMD}$$\sim$40-150\,Myr.

@@

Desite uncertainties in the details, what can be stated with
confidence is that lithium is not abundant in the spectrum of
TIC~141146667.  Figure~\ref{fig:liew_population} compares our upper
limit against the lithium equivalent width measurements reported by
\cite{Jeffries2023} based on the Gaia-ESO spectroscopic survey.  If
the star were $\lesssim$20\,Myr old, given its temperature we would
expect to see lithium in abundance ($>$400\,m\AA).  Since we do not,
we conclude that $t_{\rm Li,emp}$$\gtrsim$20\,Myr.  The
\cite{Feiden2016} isochrones provide a point for theoretical
comparison, suggesting that since $M_{\rm K}$$=$6.67\,mag, $t_{\rm
Li,th}$$\gtrsim$35\,Myr is the theoretical age at which complete
depletion occurs in a star with this luminosity (see e.g.~Figure 7
from \cite{Wood2023}).

\begin{figure}[!t]
  \centering
  \includegraphics[width=0.7\textwidth]{figures/sf3.pdf}
  \caption{Dereddened Gaia DR3 color vs.~absolute magnitude diagram of
  TIC~141146667 and comparison samples. 
  TIC~141146667 is on the pre-main sequence; stars with the same color
  on the main sequence are $\approx$1.5\,magnitudes
  fainter.  The star's location in this diagram broadly suggests an
  age of 40-150\,Myr.  }
  \label{fig:camd}
\end{figure}

@@

{\it Age: Summary}---The main indicators for the youth of
TIC~141146667 are {\it i)} that it is a complex periodic variable, and
{\it ii)} that it is 1.5 magnitudes brighter than main sequence stars
of the same color, while showing no indicators for binarity.  The
first item constitutes evidence for youth since all known CPVs are
$\lesssim$200\,Myr \cite{Bouma2024}.  The reference sequences for the
Pleiades (112\,Myr, \cite{Dahm2015}) show a few stars of equal
luminosity and the same temperature, suggesting a photometric
isochronal age upper limit $\lesssim$150\,Myr.  The weak lithium
absorption suggests an age of at least 20\,Myr based on an empirical
comparison using Gaia-ESO spectra, or at least 35\,Myr based on the
\cite{Feiden2016} isochrones.  These considerations yield our adopted
age range of 40-150\,Myr.
